---
title: "Lab 02"
author: ' Wejdan Alshateri '
subtitle: "CS3172-1, Spring 2023, Effat University"
output:
  word_document: default
  pdf_document: default
---

# Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(scales)
library(ggplot2)
``` 

# Data

```{r read-in-data, message = FALSE}
cas <- read_rds("data/canada_survey.rds")
```

# Tasks

## Task 1
Subset `cas` so that variables `energy_expense` and  `household_income` only contain values greater than 0. Overwrite  `cas`

```{r}
cas <- filter(cas,energy_expense >0, household_income >0)
```

Use function `factor()` to change the variable `marital_status` to  be a factor rather than double. Overwrite `cas`. Consult the data  dictionary and write-out what the marital status codes mean.

```{r}
cas <- mutate(cas, marital_status = factor(marital_status))
```
1. Married - a person who is legally married and living with their spouse.
2. Widowed - a person whose spouse has died and who has not remarried.
3. Divorced - a person who has been legally divorced and has not remarried.
4. Separated - a person who is legally separated from their spouse, but not divorced.
5. Never married/single - a person who has never been married, or not currently married.
6. Unknown - a person whose marital status is unknown or not reported.

## Task 2

Recode `heat_equip` so instead of having values 1, 2, 3, 4, it contains values `"steam"`, `"forced air"`, `"stove"`, and `"electric heating"` according to the data dictionary. These new values are as defined below:
o	`steam`: steam or water furnace
o	`forced air`: forced air furnace
o	`stove`: heating stoves, cookstove, or other
o	`electric heating`: electric

```{r}
cas <- mutate(cas, heat_equip = case_when(heat_equip == 1 ~ 'steam',
                                          heat_equip == 2 ~ 'forced air',
                                          heat_equip == 3 ~ 'stove',
                                          heat_equip == 4 ~ 'electric heating'))
```

Recode `heat_fuel` so instead of having values 1, 2, 3, 4, it contains values `"oil"`, `"gas"`, `"electricity"`, and `"other"` according to the data dictionary. These new values are as defined below:
o	`oil`: oil or other liquid fuel
o	`gas`: natural gas
o	`electricity`: electricity
o	`other`: bottled gas, wood, or other

```{r}
cas <- mutate(cas, heat_fuel = case_when(heat_fuel == 1 ~ 'oil',
                                         heat_fuel == 2 ~ 'gas',
                                         heat_fuel == 3 ~ 'electricity',
                                         heat_fuel == 4 ~ 'other'))
```

## Task 3

For each combination of heating fuel type and heating equipment, find the mean, median, and standard deviation of household energy expenditures. Print your results.

```{r}
cas %>% group_by(heat_equip,heat_fuel) %>% summarise(
  mean_ener_exp = mean(energy_expense),
  median_ener_exp = median(energy_expense),
  sd_ener_exp = sd(energy_expense))
```

o	What combination of fuel type and equipment has the highest average energy expense? the combination of heating equipment ( forced air) and heat fuel (oil) has the highest average energy expense equal to 3498.850.

o	Which combination has the most variability with regards to energy expense?
the combination of heating equipment ( steam ) and heat fuel (other) will have the highest standard deviation and will have the most variability in energy expense.

o	Which type of heating equipment doesn’t take all possible fuel types?
electric heating.

## Task 4

Create a bar chart of energy expenses by heating fuel type and faceted by the type of heating equipment in a 2 x 2 grid. Your axis should be appropriately labeled with a dollar sign and commas. The `scales` package may be helpful here

```{r}
ggplot(cas, mapping=aes(x=heat_fuel, y=energy_expense)) +
  geom_boxplot() + 
  scale_y_continuous(labels = scales::dollar_format()) +
  facet_wrap(~ heat_equip, nrow =2) +
  labs(title = "Energy Expense vs Heating Type",
       x = "heat fuel type",
       y = "energy expense")
```

## Task 5

Create a new variable describing the proportion of household income spent on energy related expenses, and then find the respondent that spent the highest proportion of their household income on energy and the respondent that spent the lowest proportion of their household income on energy. End your pipeline with the tibble being passed into `glimpse()`. Describe these respondents based on the data they have provided

```{r}
cas1 <- cas %>% mutate(energy_prop = energy_expense/household_income) %>%
  arrange(desc(energy_prop)) %>%
  slice(1,n()) %>%
  glimpse()
```

new column energy_prop is created, Energy_prop had a value of almost 3.78 for the respondent who spent the largest percentage of their household income on energy, This indicates the respondent is probably spending an excessive portion of their income to energy, we can see that this respondent is divorced, living in detached house, He has a relatively low household income, and he reported using electric heating and having a high level of energy consumption. On the other hand, the respondent who spent the lowest proportion of their household income on energy had energy_prop value of 1.49, which is much lower. This suggests that this respondent is using energy-efficient appliances or living in a small, energy-efficient apartment. he has a high household income of $67000 and reported a low level of consumption.


## Task 6

For each year, find the province with the cheapest median energy expense per room. Your answer should consist of a single `dplyr` pipeline that results in two rows and three columns

```{r message=FALSE}
cas %>% 
  group_by(year, province) %>% 
  summarize(median_energy_expense_per_room = median(energy_expense/rooms)) %>%
    group_by(year) %>% 
    slice(which.min(median_energy_expense_per_room))
```

## Task 7

A respondent is considered to be “energy vulnerable” if they spend more than 5% of their household income on energy expenses. Recreate the plot below, which shows the proportion of respondents who are energy vulnerable for each combination of age and education. In 2 - 3 sentences, describe what you observe in the plot.

```{r fig.width=7, fig.height=6, message=FALSE}
cas %>% mutate(energy_prop = energy_expense / household_income, 
  vulnerable = if_else(energy_prop > 0.05, "vulnerable", "not")) %>%
  group_by(education, age) %>%
  summarize(prop_vulnerable = mean(vulnerable == "vulnerable")) %>%
  ungroup() %>%
  
ggplot(aes(x = education, y = age, fill = prop_vulnerable)) + 
  geom_raster() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Education", y = "Age",
       fill = "Proportion\nenergy\nvulnerable") +
  theme_bw()
```
The graph shows that the proportion of energy-vulnerable susceptible people is higher for older people, older respondents are more likely to be energy vulnerable. and we observe a higher proportions of vulnerable households among those with lower levels of education.
